---
title: "Metabolic Flux Dynamics in Glycolysis with `dynafluxr`"
output: rmarkdown::html_vignette
author:
  - Serguei Sokol^[TBI, Université de Toulouse, CNRS, INRAE, INSA, Toulouse, France]
date: 2022-11-08
vignette: >
  %\VignetteIndexEntry{glyco}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=7,
  fig.height=5
)
```

```{r setup}
library(dynafluxr)
library(bspline)
```
## Introduction

This vignette shows how `dynafluxr` R package can be used to unravel metabolic flux dynamics from metabolite time-series measurements and a stoichiometric model. No regulation model (like Michaelis-Menten) is required for this task. Our method is heavily relying on B-splines as implemented in `bspline` package. We show how to prepare measurement data, edit stoichiometric reactions, use various package options and interpret/exploit the results. The data used for this vignette were obtained by NMR measurements on MetaToul-FluxoMet platform, Toulouse Biotechnology Institute (TBI), Toulouse, France. Credits for making these data available go to Pauline Rouane, Cyril Charlier, Guy Lippens. Pierre Millard has shared stoichiometric model. All cited persons are from TBI.

## Data preparations

Data must be stored in a `.tsv` (Tabulation Separated Values) file (`.tsv` extension is not mandatory, it can be `.txt`, `.csv`, whatever, but the field separator must be the tabulation character), one column per chemical specie. The first row describes column names. Time points, at which measurements are done, must be stored in a column whose name starts with 'Time'. No specie name can start with 'Time'. Decimal separator must be a point `.` character. A head of data file used as an example is looking like

```{r}
fmeas=system.file("dataglyco/data_PRO.tsv", package="dynafluxr")
cat(head(readLines(fmeas)), sep="\n")
```

The raw data may be difficult to read because of tab misalignment and row wrapping. Here are parsed data:
```{r}
mf=read.delim(fmeas, comment.char="#")
print(head(mf))
```
## Stoichiometric model preparation

The reactions are written as in following example:
```{r}
fsto=system.file("dataglyco/network_PRO.txt", package="dynafluxr")
cat(head(readLines(fsto)), sep="\n")
```
The first field separated by tabulation is reaction name then reaction itself where species are separated by " + " sign and possibly preceded by a stoichiometric coefficient with " * " sign.

We start by gathering example files in a freshly created working directory, say `glyco/`:
```{r}
dir.create("glyco")
file.copy(c(fmeas, fsto), "glyco/")
```
## Data exploring

Let fit available data with B-splines of order 4 and with 5 internal knots (default in `dynafluxr`), then plot them to have an idea how data looks like:
```{r}
fmeas="glyco/data_PRO.tsv"
mf=read.delim(fmeas, comment.char="#")
nm=colnames(mf)[-1L]
msp=fitsmbsp(mf$Time, mf[, nm], n=4, nki=5)
matplot(mf$Time, msp(mf$Time), t="l", ylab="Concentration [mM]", xlab="Time [s]")
matpoints(mf$Time, mf[, nm], pch="o", cex=0.5)
legend("topright", legend=nm, lty=1:5, col=1:6, cex=0.75)
```

We see that glucose (GLC) is going too close to 0, maybe even getting negative. A closer look confirms this (negative values are plotted in blue):
```{r}
ineg=msp(mf$Time, "GLC") < 0
matplot(mf$Time[!ineg], msp(mf$Time, "GLC")[!ineg], t="l", ylab="C", xlab="T", main="GLC")
abline(h=0)
matplot(mf$Time[ineg], msp(mf$Time, "GLC")[ineg], t="l", col="blue", lwd=2, add=TRUE)
matpoints(mf$Time, mf[,"GLC"], pch="o", cex=0.5)
```

As concentrations cannot get negative values by definition, we constraint B-splines to non negative values with option `positive=1`
```{r}
msp=fitsmbsp(mf$Time, mf[,-1], n=4, nki=5, positive=1)
ineg=msp(mf$Time, "GLC") < 0
print(sum(ineg))
matplot(mf$Time, msp(mf$Time, "GLC"), t="l", ylab="C", xlab="T", main="GLC")
abline(h=0)
matpoints(mf$Time, mf[,"GLC"], pch="o", cex=0.5)
```

Negative values have disappeared but at the end of time interval, around 300 s, GLC curve is increasing which has no physical meaning. Glucose is only consumed in this experimental setup. Fortunately, we can constraint B-splines to be monotonously decreasing for GLC:

```{r, fig.width=6, fig.height=4}
vmono=rep(0, length(nm))
vmono[match("GLC",nm)]=-1
msp=fitsmbsp(mf$Time, mf[,-1], n=4, nki=5, positive = 1, monotone = vmono)
matplot(mf$Time, msp(mf$Time, "GLC"), t="l", ylab="C", xlab="T", main="GLC")
matpoints(mf$Time, mf[,"GLC"], pch="o", cex=0.5)
```

Looks good now. In `dynafluxr`, we have no option to constraint B-splines to non negative values, because it is done automatically. On the other hand, we do need to indicate which specie is supposed to be monotonously decreasing or increasing. It can be be done with `--mono MONO` option where `MONO` is a file name appropriately formatted to indicate which metabolite is increasing and which is decreasing. It can also be done with options `--incresing` and `--decreasing` with corresponding coma separated lists of specie names. It is this last option that we will use with `dynafluxr` but for now, let check some other metabolites, e.g. pyruvate (PYR) to be sure that the fit looks good too:
```{r, fig.width=6, fig.height=4}
matplot(mf$Time, msp(mf$Time, "PYR"), t="l", ylab="C", xlab="T", main="PYR")
matpoints(mf$Time, mf[,"PYR"], pch="o", cex=0.5)
```

Looks OK, so we are set to move to the next step.

## First estimate of fluxes with `dynafluxr`

We try `dynafluxr::cli()` (Command Line Interface) function with minimal options, just indicating measurement data, stoichiometric model and, as seen before, that GLC must be monotonously decreasing (`--decreasing` is abbreviated to `--decr`):

```{r}
res=cli(c("-m", "glyco/data_PRO.tsv", "-s", "glyco/network_PRO.txt", "--decr", "GLC"))
```

The same command could be run from a shell (not R) environment. In this case it would look like:

```
$ Rscript -e "dynafluxr::cli()" -m glyco/data_PRO.tsv -s glyco/network_PRO.txt --decr GLC
```

The full list of available options can be viewed with `dynafluxr::cli("-h")` command. The result of `cli()` function (in case of no error in run-time), is a series of data and pdf files written in a directory `glyco/data_PRO` which is the stump of measurement data file:

```{r}
list.files("glyco/data_PRO/")
```

The file `Readme.md` describes the content of the result directory. These files can be explored by system tools like pdf-viewers and spreadsheet software but for needs of this vignette we will reproduce some results by R means. As we won't need result files in this demo page, we'll cancel their writing with empty option `-o ""` which normally indicates result directory/archive name if it is different from the default one:
```{r}
res=cli(c("-m", "glyco/data_PRO.tsv", "-s", "glyco/network_PRO.txt", "--decr", "GLC", "-o", ""))
```

Let now glance on fit quality, i.e. results of $\chi^2$-test:
```{r}
print(res$chi2tab)
```

We can see that all $p$-values are 0 which means a very poor fit quality. Let see what is going on by plotting integrated species with measurements:

```{r}
nm=colnames(res$mf)[-1]
matplot(res$tpp, res$isp(res$tpp, nm), t="l", xlab="T", ylab="C")
matpoints(res$tp, res$mf[,nm], pch="o", cex=0.5)
legend("topright", legend=nm, lty=1:5, col=1:6, cex=0.75)
```

Indeed, for all species, integrated curves are not fitting well the corresponding data. We take a closer look on pyruvate (PYR) which is buried in the full graph:
```{r}
plot(res$tpp, res$isp(res$tpp, "PYR"), t="l", xlab="T", ylab="C", ylim=range(res$mf$PYR))
points(res$tp, res$mf$PYR, pch="o", cex=0.5)
```
We will use PYR example for tracking the fit quality as fit options change.

## Getting fit better

Let see the integral residual of metabolites which are not measured and thus are supposed to be close to 0:

```{r}
nmall=rownames(res$sto)
nm=colnames(res$mf)[-1L]
nm0=setdiff(nmall, nm)
matplot(res$tpp, res$risp(res$tpp, nm0), t="l", ylab="M-\u222bSv dt", xlab="T")
legend("topright", legend=nm0, lty=1:5, col=1:6, cex=0.75)
```

We can see that the most diverging metabolites are fructose bi-phosphate (FBP) and fructose-6-phosphate (F6P). We can try to declare them as not available (NA) thus they won't degrade least-squares objective function with their diverging from 0 residuals:
```{r}
res=cli(c("-m", "glyco/data_PRO.tsv", "-s", "glyco/network_PRO.txt", "--decr", "GLC", "-o", "", "--lna", "FBP,F6P"))
nmall=rownames(res$sto)
nm=colnames(res$mf)[-1L]
nm0=setdiff(nmall, nm)
layout(t(1:2))
matplot(res$tpp, res$risp(res$tpp, nm0), t="l", ylab="M-\u222bSv dt", xlab="T")
legend("topright", legend=nm0, lty=1:5, col=1:6, cex=0.75)
plot(res$tpp, res$isp(res$tpp, "PYR"), t="l", xlab="T", ylab="C", ylim=range(res$mf$PYR), main="PYR")
points(res$tp, res$mf$PYR, pch="o", cex=0.5)
```
There is some progress in fit quality but PYR is still not well fitted. We try to add next diverging metabolites to NA: BPG (a little bit buried in the plot) and ACCOA.
```{r}
res=cli(c("-m", "glyco/data_PRO.tsv", "-s", "glyco/network_PRO.txt", "--decr", "GLC", "-o", "", "--lna", "FBP,F6P,BPG,ACCOA"))
nmall=rownames(res$sto)
nm=colnames(res$mf)[-1L]
nm0=setdiff(nmall, nm)
layout(t(1:2))
matplot(res$tpp, res$risp(res$tpp, nm0), t="l", ylab="M-\u222bSv dt", xlab="T")
legend("topright", legend=nm0, lty=1:5, col=1:6, cex=0.75)
plot(res$tpp, res$isp(res$tpp, "PYR"), t="l", xlab="T", ylab="C", ylim=range(res$mf$PYR), main="PYR")
points(res$tp, res$mf$PYR, pch="o", cex=0.5)
```

Now, residuals for "0" metabolites are all very close to 0 which means that they are not degrading least squares anymore and PYR is very well fitted. A good overall fit is confirmed by $\chi^2$-test:
```{r}
print(res$chi2tab)
```

as all $p$-values are well above 0.05 usual threshold.
We are ready to see the main result: estimated reaction rates.
```{r}
matplot(res$tpp, res$vsp(res$tpp), t="l", ylab="Rate [1/s]", xlab="T")
legend("topright", colnames(bsppar(res$vsp)$qw), lty=1:5, col=1:6, cex=0.75)
```
It can be observed some oscillations in reaction rates and we must check if they don't come from a possible overfitting.

## Exploring problem of overfitting

By default, the number of internal knots used for fitting is 5. Let few values under and above (1 to 6) and see if the measurements are still well fitted. We will continue to use PYR for this purpose:
```{r}
layout(matrix(1:6, nrow=2, byrow=TRUE))
resk=lapply(1:6, function(k) {
  r=cli(c("-m", "glyco/data_PRO.tsv", "-s", "glyco/network_PRO.txt", "--decr", "GLC", "-o", "", "--lna", "FBP,F6P,BPG,ACCOA", "-k", k))
  plot(r$tpp, r$isp(r$tpp, "PYR"), main=paste0("PYR k=",k,sep=""), ylab="C", xlab="T", t="l")
  points(r$tp, r$mf$PYR, pch="o", cex=0.5)
  r
})
```
Starting from value `k=3`, the fit looks satisfactory. Let see what says $chi^2$ test:
```{r}
for (r in resk)
  print(r$chi2tab)
```
We can see that $p$-value is above 0.05 for all metabolites only for k=5. We will retain this value as the given the best fit.

The total data fit for k=5 looks like:
```{r}
res=resk[[5L]]
nm=colnames(res$mf)[-1L]
matplot(res$tpp, res$isp(res$tpp, nm), t="l")
matpoints(res$tp, res$mf[,nm], pch="o", cex=0.5)
legend("topright", legend=nm, lty=1:5, col=1:6)
```
and reaction rates:
```{r}
matplot(res$tpp, res$vsp(res$tpp), t="l", ylab="Rate [1/s]", xlab="T")
legend("topright", colnames(bsppar(res$vsp)$qw), lty=1:5, col=1:6, cex=0.75)
```
The curve oscillations are present for some fluxes but as seen before, in this case, it is not due to overfitting but reflects system behavior through measured data.

In the result files, the rates are plotted with semi-transparent error bands ±2·SD. Let do it here too:
```{r}
nmreac=colnames(bsppar(res$vsp)$qw)
matplot(res$tpp, res$vsp(res$tpp), t="l", ylab="Rate [1/s]", xlab="T")
tppr=c(res$tpp, rev(res$tpp))
ireac=0
for (reac in nmreac) {
  vp=res$vsp(res$tpp, reac, fsd=2)
  vm=rev(res$vsp(res$tpp, reac, fsd=-2))
  polygon(tppr, c(vp, vm), border=NA, col=do.call(rgb, as.list(c(col2rgb(ireac%%6+1)/255, 0.3))))
  ireac=ireac+1
}
legend("topright", nmreac, lty=1:5, col=1:6, cex=0.75)
```

Having estimated rates, allow us to explore details of metabolic fluxes. We can have not only the total flux $\frac{dM_i}{dt}$ but all its comonents coming from involved reactions $S_{ij}v_j$. For PYR, it gives:
```{r}
nm="PYR"
jnz=names(which(res$stofull[nm,] != 0)) # reactions involved in PYR mass balance
fl=t(res$stofull[nm,jnz]*t(res$vsp(res$tpp, jnz)))
fl=cbind(res$fsp(res$tpp, nm), fl)
matplot(res$tpp, fl, t="l", ylab="Flux [mM/s]", xlab="Time [s]", main="PYR flux detail")
legend("topright", legend=c("Total", jnz), lty=1:5, col=1:6)
```

## Conclusion

In this vignette, we have shown a step-by-step procedure for reaction rate estimation from time-series metabolite measurements and a stoichiometric model. We have seen how package parameters can be chosen, how fit quality can be evaluated and how data can be presented and exploited. Based on this vignette, user should be ready to work with his own data set and model.
